\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-12 paper.qmd
Finalize the model for S-curve data

The performance and robustness of our model depend on three key parameters: (i) the total number of bins ($b$), (ii) a benchmark value used to remove low-density hexagons, and (iii) a benchmark value used to remove long edges. However, there is no analytical formula to calculate an appropriate value for these parameters (see Appendix for details on how default values are calculated). The selection of these parameter values depends on the model performance computed by Mean Squared Error (MSE) (see @sec-goodfit).

@@

Furthermore, the total number of bins is determined by the number of bins along the x-axis and y-axis (see @eq-equation13). To calculate the effective total number of bins, the minimum and approximate maximum number of bins along the x-axis are used as candidate values. The minimum number of bins along the x-axis is $1$. The maximum number of bins along the x-axis can be estimated by taking the square root of the number of NLDR data points. To investigate the impact of the number of bins on MSE, the analysis is focused on the MSE for different total numbers of bins. This range encompasses the total number of bins calculated for the minimum and maximum values of the number of bins along the x-axis.

Typically, the MSE plot will have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope will gradually decline or level off, indicating that a higher number of bins generates a smaller error. This MSE plot is useful in determining the effective number of bins required to construct the 2D model. @fig-diagnosticpltScurve (b) has created a graph that shows how the Mean Squared Error (MSE) changes depending on the number of bins assigned in the hexagonal grid for various NLDR techniques applied to the S-curve data set. As per the rule of thumb mentioned earlier, the effective number of bins for each NLDR method is shown in the @tbl-msebinsnldr. Furthermore, @fig-modelScurve shows the 2D models constructed for each NLDR technique with the selected total number of bins and how the models look in high-D with original data.

@@

```{r}
#| eval: false
#| echo: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset. What is the effective number of bins in each NLDR technique to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to 144, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-diagnosticpltScurve
##| out-width: 100%
#| fig-pos: H

@@

@@

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset. What is the effective number of bins in each NLDR technique to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to 144, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-diagnosticpltPBMC
##| out-width: 100%
#| fig-pos: H

@@

result1_2d_layout <- df_res_1_2_label |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = result1_class_label))+
    geom_point(alpha=0.5) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#d82cd3", "#49df20", "#619cff", "#000000", "#df2020", "#ffbd61", "yellow")) +
  # geom_text(aes(x=UMAP1,y = UMAP2,label = cell_label), data = class_avg,inherit.aes = F, color = "black",fontface = "bold",size = 2) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

result2_2d_layout <- df_res_1_2_label |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = result2_class_label))+
    geom_point(alpha=0.5) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#d82cd3", "#49df20", "#619cff", "#000000", "#df2020", "#ffbd61", "yellow")) +
  # geom_text(aes(x=UMAP1,y = UMAP2,label = cell_label), data = class_avg,inherit.aes = F, color = "black",fontface = "bold",size = 2) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

result1_2d_layout + result2_2d_layout
```

@@

# show_langevitour(df_all_mnist, df_bin_mnist, df_bin_centroids_mnist, benchmark_mnist, distance_mnist, "distance", col_start = "PC")
```

>>>>

The performance and robustness of our model depend on four key parameters: (i) the total number of bins ($b$), (ii) a benchmark value used to remove low-density hexagons, (iii) a benchmark value used to remove long edges, and (iv) starting point of the hexagonal grid. However, there is no analytical formula to calculate an appropriate value for these parameters (the computation of default values can be find in *quollr* paper). The selection of these parameter values depends on the model performance computed by Mean Squared Error (MSE) (see @sec-goodfit).

@@

$$
b = b_1 \times b_2
$${#eq-equation8}

Furthermore, the total number of bins is determined by the number of bins along the x-axis and y-axis (see @eq-equation8). To calculate the effective total number of bins, candidate values are selected based on the range between the minimum and approximate maximum number of bins along the x and y axes. The minimum number of bins along each axis is set to $1$, while the maximum number is estimated by taking the square root of the NLDR data point count. The analysis evaluates the MSE across varying total bin counts within this range, covering the minimum to maximum values along both axes.

The analysis focuses on exploring the effect of total number of bins on MSE. It considers MSE across different total number of bins, covering the range calculated from the minimum to maximum values along both x and y axes.

Typically, the MSE plot tends to have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or levesl off, indicating that a higher number of bins generates a smaller error. This MSE plot is useful in determining the effective number of bins required to construct the 2D model. @fig-diagnosticpltScurve (b) has created a graph that shows how the Mean Squared Error (MSE) changes depending on the number of bins assigned in the hexagonal grid for UMAP applied to the S-curve dataset. As per the rule of thumb mentioned earlier, the effective number of bins is determined to be $84$, with $6$ bins along the x-axis and $14$ bins along the y-axis. Furthermore, @fig-modelScurve shows the 2D model constructed using UMAP with the selected total number of bins and how the model overlay on high-D data.

umap_s_curve_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = UMAP_data, 
                                    x = "UMAP1", y = "UMAP2"))) |>
  dplyr::rename(c("UMAP1" = "scaled_UMAP1", 
                  "UMAP2" = "scaled_UMAP2")) |>
  dplyr::mutate(ID = 1:NROW(UMAP_data))

hex_size_vec <- seq(0.06, 2, by = 0.01)

mse_df <- dplyr::bind_rows(vec)[0, ]
mse_df <- mse_df |>
  dplyr::mutate_if(is.character, as.numeric)

  num_bin_list <- calc_bins(data = umap_s_curve_scaled, 
            x = "UMAP1", y = "UMAP2", 
            hex_size = hex_size_vec[i], buffer_x = NA, buffer_y = NA)

  model_object <- fit_highd_model( training_data = training_data, 
                                   nldr_df_with_id = umap_s_curve_scaled, 
                                   x = "UMAP1", y = "UMAP2", 
                                   num_bins_x = num_bins_x, 
                                   num_bins_y = num_bins_y, 
                                   x_start = NA, y_start = NA, 
                                   buffer_x = NA, buffer_y = NA, 
                                   hex_size = hex_size_vec[i],
                                   is_rm_lwd_hex = FALSE, 
                                   benchmark_to_rm_lwd_hex = NA, 
                                   col_start_2d = "UMAP", 
                                   col_start_highd = "x")

  pred_emb_list <- predict_emb(test_data = training_data, 
                                  df_bin_centroids = centroid_df_training, 
                                  df_bin = avg_df_training, type_NLDR = "UMAP")

  eval_list <- gen_summary(test_data = training_data, 
                                  prediction_df = pred_df_training, 
                                  df_bin = avg_df_training, col_start = "x")

  mse_df <- mse_df |>
    tibble::add_row(num_bins = num_bins_x * num_bins_y,
                    aic = eval_list$aic,
                    mse = eval_list$mse,
                    num_bins_x = num_bins_x,
                    num_bins_y = num_bins_y,
                    hex_size = hex_size_vec[i])

## If same total number of bins occurred only select ones with minimum error
### Obtain duplicate bins
dupli_bins <- mse_df |> 
  dplyr::count(num_bins) |> 
  dplyr::filter(n > 1) |> 
  dplyr::pull(num_bins)

### Group split by duplicated bins
duplicate_df_list <- mse_df |>
  dplyr::filter(num_bins %in% dupli_bins) |>
  dplyr::arrange(num_bins) |>
  dplyr::group_split(num_bins)

### Obtain the mse_df with not duplicated bins
not_dupli_df <- mse_df |>
  dplyr::filter(!(num_bins %in% dupli_bins))

### Combine duplicated and not duplicated(corrected) bins dfs
mse_df <- dplyr::bind_rows(not_dupli_df, duplicate_df)
```

## To draw with AIC
aic_plot <- ggplot(mse_df, aes(x = num_bins, y = aic
)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 84, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  #scale_colour_manual(values = c("#377eb8")) +
  ylab("AIC") +
  xlab("total number of bins")
## Effective number of bins along x-axis

mse_plot <- ggplot(mse_df, aes(x = num_bins,
                                       y = mse
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 84, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  #scale_colour_manual(values = c("#377eb8")) +
  ylab("MSE") +
  xlab("total number of bins")

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset. What is the effective number of bins in each NLDR technique to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to $84$, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-diagnosticpltScurve
##| out-width: 100%
#| fig-pos: H

aic_plot + mse_plot +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

@@

@@

## Decide by looking at MSE plot
num_bins_x_s_curve <- 6
num_bins_y_s_curve <- 14
hex_size_s_curve <- 0.13

hb_obj_s_curve <- hex_binning(data = umap_s_curve_scaled, x = "UMAP1", 
                      y = "UMAP2", num_bins_x = num_bins_x_s_curve, 
                      num_bins_y = num_bins_y_s_curve, x_start = NA, y_start = NA, 
                      buffer_x = NA, buffer_y = NA, hex_size = hex_size_s_curve, col_start = "UMAP")

## Data set with all possible centroids in the hexagonal grid
all_centroids_df <- as.data.frame(do.call(cbind, hb_obj_s_curve$centroids))

## Generate all coordinates of hexagons
hex_grid <- as.data.frame(do.call(cbind, hb_obj_s_curve$hex_poly))

## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_s_curve$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
pcentroid_plot_s_curve <- ggplot(data = hex_grid, aes(x = x, y = y)) +
  geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
  geom_point(data = df_bin_centroids, aes(x = c_x, y = c_y), color = "#33a02c") +
  coord_fixed() +
  theme_linedraw() +
  theme(legend.position = "bottom", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

umap_data_with_hb_id_pbmc <- as.data.frame(do.call(cbind, hb_obj_s_curve$data_hb_id))

model_object <- fit_highd_model( training_data = training_data, 
                                 nldr_df_with_id = umap_s_curve_scaled, 
                                 x = "UMAP1", y = "UMAP2", 
                                 num_bins_x = num_bins_x_s_curve, 
                                 num_bins_y = num_bins_y_s_curve, 
                                 x_start = NA, y_start = NA, 
                                 buffer_x = NA, buffer_y = NA, 
                                 hex_size = hex_size_s_curve,
                                 is_rm_lwd_hex = FALSE, 
                                 benchmark_to_rm_lwd_hex = NA, 
                                 col_start_2d = "UMAP", 
                                 col_start_highd = "x")

df_bin_centroids_s_curve <- model_object$df_bin_centroids
df_bin_s_curve <- model_object$df_bin

## Triangulate bin centroids
tr1_object_s_curve <- tri_bin_centroids(df_bin_centroids_s_curve, x = "c_x", y = "c_y")
tr_from_to_df_s_curve <- gen_edges(tri_object = tr1_object_s_curve)

## Compute 2D distances
distance_s_curve <- cal_2d_dist(tr_coord_df = tr_from_to_df_s_curve, 
                             start_x = "x_from", start_y = "y_from", 
                             end_x = "x_to", end_y = "y_to", 
                             select_vars = c("from", "to", "distance"))

## To find the benchmark value
benchmark_s_curve <- find_lg_benchmark(distance_edges = distance_s_curve, distance_col = "distance")

trimesh_s_curve_umap <- vis_lg_mesh(distance_edges = distance_s_curve, benchmark_value = benchmark_s_curve, tr_coord_df = tr_from_to_df_s_curve, distance_col = "distance")

trimesh_s_curve_umap <- trimesh_s_curve_umap +
  theme_linedraw() +
  theme(legend.position = "bottom", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3) +
  labs(colour = "")

trimesh_removed_s_curve_umap <- vis_rmlg_mesh(distance_edges = distance_s_curve, benchmark_value = benchmark_s_curve, tr_coord_df = tr_from_to_df_s_curve, distance_col = "distance")

trimesh_removed_s_curve_umap <- trimesh_removed_s_curve_umap +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'c', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)
```

```{r}
#| echo: false
#| fig-cap: "(a), (b) 2D model."
#| label: fig-modelScurve

pcentroid_plot_s_curve + trimesh_s_curve_umap + trimesh_removed_s_curve_umap +
  plot_layout(guides='collect', ncol = 3) &
  theme(legend.position='bottom') 
```

@@

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training PBMC3k dataset. What is the effective number of bins in each NLDR technique to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to 80, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-diagnosticpltPBMC
##| out-width: 100%
#| fig-pos: H

@@

result1_2d_layout <- df_res_1_2_label |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = result1_class_label))+
    geom_point(alpha=0.5) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#d82cd3", "#49df20", "#619cff", "#000000", "#df2020", "#ffbd61", "yellow")) +
  # geom_text(aes(x=UMAP1,y = UMAP2,label = cell_label), data = class_avg,inherit.aes = F, color = "black",fontface = "bold",size = 2) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3) +
  labs(colour = "cluster labels")

result2_2d_layout <- df_res_1_2_label |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = result2_class_label))+
    geom_point(alpha=0.5) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(legend.position = "bottom", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#d82cd3", "#49df20", "#619cff", "#000000", "#df2020", "#ffbd61", "yellow")) +
  # geom_text(aes(x=UMAP1,y = UMAP2,label = cell_label), data = class_avg,inherit.aes = F, color = "black",fontface = "bold",size = 2) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3) +
  labs(colour = "cluster labels")

@@

```{r}
#| echo: false
#| fig-cap: "2D layout from UMAP (n_neighbors = 5, min_dist = 0.99, metric = cosine) applied for PBMC3k dataset (a) colored by manualy identified clusters, (b) colored by `detour` results"
#| label: fig-nldervisPBMCUMAP3Res
#| out-width: 100%

result1_2d_layout + result2_2d_layout +
  plot_layout(guides='collect', ncol=2) &
  theme(legend.position='bottom')

@@

# show_langevitour(df_all_mnist, df_bin_mnist, df_bin_centroids_mnist, benchmark_mnist, distance_mnist, "distance", col_start = "PC")

predict_minst_df <- as.data.frame(do.call(cbind, predict_emb(training_data_mnist, df_bin_centroids_mnist, df_bin_mnist, "PaCMAP")))
gen_summary(training_data_mnist, predict_minst_df, df_bin_mnist, col_start = "PC")
```

### 2024-03-12 paper.tex
Finalize the model for S-curve data

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable]}{\end{tcolorbox}}\fi

@@

The performance and robustness of our model depend on three key
parameters: (i) the total number of bins (\(b\)), (ii) a benchmark value
used to remove low-density hexagons, and (iii) a benchmark value used to
remove long edges. However, there is no analytical formula to calculate
an appropriate value for these parameters (see Appendix for details on
how default values are calculated). The selection of these parameter
values depends on the model performance computed by Mean Squared Error
(MSE) (see Section~\ref{sec-goodfit}).

@@

Furthermore, the total number of bins is determined by the number of
bins along the x-axis and y-axis (see \textbf{?@eq-equation13}). To
calculate the effective total number of bins, the minimum and
approximate maximum number of bins along the x-axis are used as
candidate values. The minimum number of bins along the x-axis is \(1\).
The maximum number of bins along the x-axis can be estimated by taking
the square root of the number of NLDR data points. To investigate the
impact of the number of bins on MSE, the analysis is focused on the MSE
for different total numbers of bins. This range encompasses the total
number of bins calculated for the minimum and maximum values of the
number of bins along the x-axis.

Typically, the MSE plot will have a steep slope at the beginning,
indicating that a smaller number of bins causes a larger amount of
error. Then, the slope will gradually decline or level off, indicating
that a higher number of bins generates a smaller error. This MSE plot is
useful in determining the effective number of bins required to construct
the 2D model. \textbf{?@fig-diagnosticpltScurve} (b) has created a graph
that shows how the Mean Squared Error (MSE) changes depending on the
number of bins assigned in the hexagonal grid for various NLDR
techniques applied to the S-curve data set. As per the rule of thumb
mentioned earlier, the effective number of bins for each NLDR method is
shown in the \textbf{?@tbl-msebinsnldr}. Furthermore,
\textbf{?@fig-modelScurve} shows the 2D models constructed for each NLDR
technique with the selected total number of bins and how the models look
in high-D with original data.

@@

Consequently, it is not necessary to remove low-density hexagons when
constructing the 2D model for the S-curve dataset in NLDR techniques,
except for PHATE. While there may be some fluctuations in MSE between
benchmark values of 0 to 1, the absence of significant changes suggests
that maintaining low-density hexagons is preferable. However, to choose
the best selection, it's recommended to examine the benchmark values
around the first local minimum in the MSE curve. However, the MSE for
PHATE fluctuates when benchmark values range between 0 and 1. Therefore,
it is necessary to explore the benchmark values surrounding the first
local minimum to determine an effective benchmark value. There is no
universal rule for selecting a process. As shown in
\textbf{?@fig-diagnosticpltScurvelwd}, the MSE values for PHATE
fluctuate, and the first local minimum is at \(0\). Therefore, nearby
benchmark values such as \(0.01\), \(0.04\), and \(0.06\) were explored
to identify the best value for preserving the data structure. In this
case, a benchmark value of \(0.06\) was chosen. The resulting 2D model
is shown in \textbf{?@fig-modelScurve} (c), while the video link
provides a visual representation of how the model appears in high-D.

@@

@@

\caption{\label{fig-diagnosticpltPBMC}Goodness of fit statistics from
UMAP applied to training S-curve dataset. What is the effective number
of bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
144, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-68-1.pdf}

@@

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, sharp corners, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden]}{\end{tcolorbox}}\fi

@@

The performance and robustness of our model depend on four key
parameters: (i) the total number of bins (\(b\)), (ii) a benchmark value
used to remove low-density hexagons, (iii) a benchmark value used to
remove long edges, and (iv) starting point of the hexagonal grid.
However, there is no analytical formula to calculate an appropriate
value for these parameters (the computation of default values can be
find in \emph{quollr} paper). The selection of these parameter values
depends on the model performance computed by Mean Squared Error (MSE)
(see Section~\ref{sec-goodfit}).

@@

\begin{equation}\protect\hypertarget{eq-equation8}{}{
b = b_1 \times b_2
}\label{eq-equation8}\end{equation}

Furthermore, the total number of bins is determined by the number of
bins along the x-axis and y-axis (see Equation~\ref{eq-equation8}). To
calculate the effective total number of bins, candidate values are
selected based on the range between the minimum and approximate maximum
number of bins along the x and y axes. The minimum number of bins along
each axis is set to \(1\), while the maximum number is estimated by
taking the square root of the NLDR data point count. The analysis
evaluates the MSE across varying total bin counts within this range,
covering the minimum to maximum values along both axes.

The analysis focuses on exploring the effect of total number of bins on
MSE. It considers MSE across different total number of bins, covering
the range calculated from the minimum to maximum values along both x and
y axes.

Typically, the MSE plot tends to have a steep slope at the beginning,
indicating that a smaller number of bins causes a larger amount of
error. Then, the slope gradually declines or levesl off, indicating that
a higher number of bins generates a smaller error. This MSE plot is
useful in determining the effective number of bins required to construct
the 2D model. Figure~\ref{fig-diagnosticpltScurve} (b) has created a
graph that shows how the Mean Squared Error (MSE) changes depending on
the number of bins assigned in the hexagonal grid for UMAP applied to
the S-curve dataset. As per the rule of thumb mentioned earlier, the
effective number of bins is determined to be \(84\), with \(6\) bins
along the x-axis and \(14\) bins along the y-axis. Furthermore,
Figure~\ref{fig-modelScurve} shows the 2D model constructed using UMAP
with the selected total number of bins and how the model overlay on
high-D data.

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltScurve-1.pdf}

\caption{\label{fig-diagnosticpltScurve}Goodness of fit statistics from
UMAP applied to training S-curve dataset. What is the effective number
of bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
\(84\), the slope of the Mean Squared Error (MSE) plot experiences a
sudden and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

@@

Consequently, it is not necessary to remove low-density hexagons when
constructing the 2D model for the S-curve dataset in NLDR techniques,
except for PHATE. While there may be some fluctuations in MSE between
benchmark values of 0 to 1, the absence of significant changes suggests
that maintaining low-density hexagons is preferable. However, to choose
the best selection, it's recommended to examine the benchmark values
around the first local minimum in the MSE curve. However, the MSE for
PHATE fluctuates when benchmark values range between 0 and 1. Therefore,
it is necessary to explore the benchmark values surrounding the first
local minimum to determine an effective benchmark value. There is no
universal rule for selecting a process. As shown in
\textbf{?@fig-diagnosticpltScurvelwd}, the MSE values for PHATE
fluctuate, and the first local minimum is at \(0\). Therefore, nearby
benchmark values such as \(0.01\), \(0.04\), and \(0.06\) were explored
to identify the best value for preserving the data structure. In this
case, a benchmark value of \(0.06\) was chosen. The resulting 2D model
is shown in Figure~\ref{fig-modelScurve} (c), while the video link
provides a visual representation of how the model appears in high-D.

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-modelScurve-1.pdf}

\caption{\label{fig-modelScurve}(a), (b) 2D model.}

@@

\caption{\label{fig-diagnosticpltPBMC}Goodness of fit statistics from
UMAP applied to training PBMC3k dataset. What is the effective number of
bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
80, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervisPBMCUMAP3Res-1.pdf}

\caption{\label{fig-nldervisPBMCUMAP3Res}2D layout from UMAP
(n\_neighbors = 5, min\_dist = 0.99, metric = cosine) applied for PBMC3k
dataset (a) colored by manualy identified clusters, (b) colored by
\texttt{detour} results}

@@

\begin{verbatim}
$mse
[1] 3.017065

$aic
[1] 87724.49
\end{verbatim}
